---
title: "Clean GDSC Dataset"
linkcolor: blue
output:
  rmdformats::readthedown:
    highlight: kate
    number_sections: true
    code_folding: show
---

```{r setup, include=FALSE}
library(knitr)

options(max.print="75")
opts_chunk$set(comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)

library(here)
library(tidyverse)
```


# Introduction

In order to run this script, the `raw_GDSC2_data.zip` file (found in the `Data` subdirectory of this project) must be unzipped.
These unzipped files are too large to be pushed to the GitHub directory unless using a paid version with large file storage. 
The original dataset (along with all the raw data) is documented at [www.cancerrxgene.org](https://www.cancerrxgene.org). 

In its entirity, this script should take 2-3 minutes to knit. 

# Meta Cell Line Data (downloaded 22 August 19)

## Read Data
Here we read in the meta data for GDSC. 
This contains all the cell lines by various identifiers (we use `COSMIC_ID`) and descriptions of each line, including tissue and histology of origin. 

```{r read_gdsc_meta}

cell_line_details <- read.csv(here("Data", "raw_GDSC2_Data", "cell_line_details.csv"), 
                               stringsAsFactors = FALSE, na.strings = c("", "UNABLE TO CLASSIFY"))

```

## Clean Data
We also do a couple quick cleaning steps:

1. Adjusting a variable name containing a forward slash (`/`) to avoid errors later.
2. Removing the row containing the total values

```{r clean_gdsc_meta}
# Adjust label for easier reading
levels(cell_line_details$Cancer.Type..matching.TCGA.label.) <- 
  c(levels(cell_line_details$Cancer.Type..matching.TCGA.label.), "COAD&READ")
cell_line_details$Cancer.Type..matching.TCGA.label.[cell_line_details$Cancer.Type..matching.TCGA.label.=="COAD/READ"] <- "COAD&READ"

# Remove row that doesn't represent a single entry
cell_line_details <- cell_line_details[!(cell_line_details$Sample.Name =="TOTAL:"), ]
```


# Drug Response Data (downloaded 22 August 19)

## Read Data

Here we upload the drug response data from the second version of the GDSC dataset (GDSC2). Each cell line can still be identified using `COSMIC_ID`.
```{r upload_gdsc_drug}
dr_gdsc2 <- read.csv(here("Data", "raw_GDSC2_Data", "GDSC2_fitted_dose_response_17Jul19.csv"), 
                     stringsAsFactors = FALSE)
```

## Clean Data
We clean the data by adjusting data types and converting IC50 from natural log to log2.

```{r clean_gdsc_drug}
dr_gdsc2$MIN_CONC <- as.numeric(dr_gdsc2$MIN_CONC)
dr_gdsc2$MAX_CONC <- as.numeric(dr_gdsc2$MAX_CONC)
dr_gdsc2$LN_IC50 <- as.numeric(dr_gdsc2$LN_IC50)
dr_gdsc2$AUC <- as.numeric(dr_gdsc2$AUC)
dr_gdsc2$RMSE <- as.numeric(dr_gdsc2$RMSE)
dr_gdsc2$Z_SCORE <- as.numeric(dr_gdsc2$Z_SCORE)
dr_gdsc2$LN_IC50 <- exp(dr_gdsc2$LN_IC50)
dr_gdsc2$LN_IC50 <- log2(dr_gdsc2$LN_IC50)
dr_gdsc2_colnames <- colnames(dr_gdsc2)
dr_gdsc2_colnames[16] <- "IC50"
colnames(dr_gdsc2) <- dr_gdsc2_colnames
```


Let's take a look at what our dataframe looks like now.

```{r examine_dr_data}
colnames(dr_gdsc2)

tbl_df(dr_gdsc2)
```


# Expression Data (downloaded 22 August 19)

## Read Data

We'll read in the GDSC expression data, which comes from microarray experiments. 
Genes that have a blank entry for the gene symbol are removed.
Details on the experimental protocols can be found [here](https://www.cancerrxgene.org/gdsc1000/GDSC1000_WebResources/Home.html). 

```{r read_gdsc_expr}
expr_gdsc_data <- read.csv(here("Data", "raw_GDSC2_Data", "Cell_line_RMA_proc_basalExp.csv"),
                           stringsAsFactors = FALSE)
expr_gdsc_data <- expr_gdsc_data[!(expr_gdsc_data$GENE_SYMBOLS==""), ]
```


## Clean Data

In order to compare to TCGA data later, we must have matching ID names for each gene. GDSC labels genes using HGNC IDs, while TCGA uses Entrez ID. 
Below, we convert GDSC gene names to Entrez ID using biomaRt. 
Because the biomaRt query can take some time to complete, the results are saved to this repository. 
If the user wishes to run the query themselves, they may uncomment all lines except the last line in the following code chunk. 

```{r query_hgnc2entrez}

# gene_hgnc_symbols <- expr_gdsc_data$GENE_SYMBOLS
# ensembl = useDataset("hsapiens_gene_ensembl",mart=useMart("ensembl"))
# gdsc_hgnc_entrez <- getBM(attributes=c('hgnc_symbol', 'entrezgene_id', 
#                                        "external_gene_name"), 
#                           filters = 'hgnc_symbol', 
#                           values = gene_hgnc_symbols, 
#                           uniqueRows = TRUE,
#                           mart = ensembl)
# saveRDS(gdsc_hgnc_entrez, "../biomaRt_query.rds")
gdsc_hgnc_entrez <- readRDS(here("biomaRt_query_091619.rds"))
```


Once we have a data structure containing both symbols, we can convert the GDSC symbols from HGNC to Entrez IDs. 
This requires removal of any genes from the GDSC set that don't map to an Entrez ID. 
It also requires manually searching through the duplicate mappings to determine which Entrez ID should be used. 
Additionally, there were two sets of two GDSC genes that mapped to the same Entrez ID. 
The best HGNC genes to keep were determined manually as well.  

```{r convert_gdsc_genenames}
# Remove symbols that don't have Entrez ID
gdsc_hgnc_entrez_trunc <- gdsc_hgnc_entrez[!(is.na(gdsc_hgnc_entrez$entrezgene_id)), ]

# Find duplicated hgnc id's
dups <- gdsc_hgnc_entrez_trunc[duplicated(gdsc_hgnc_entrez_trunc$hgnc_symbol), "hgnc_symbol"]
remove_genes <- c("2844", "111089941", "107987478", "107987479", "100037417", 
             "102724788", "107984125", "56260", "100533105", "6013", 
             "101929777", "90632", "5554", "107986084", "104909134", "266",
             "107080644", "100132596", "100913187", "3742", "643314",
             "100653049", "84953", "100533467", "106821730", "9103", "388289",
             "107987457", "107080638", "388813", "109504726", "102723899",
             "100528062", "105372824", "110116772", "100534012", "285464", 
             "112267859", "105372240", "113455421", "348738", "80111",
             "102724985", "105374836", "110354863", "23285", "392490", 
             "162967", "100526832", "107987373", "286223", "51326", 
             "100996709", "267004", "3117", "653166", "100532736") # Manually determined from searching duplicate entrez ids 
remove_rows <- c("15619", "14482") # Manually determined which should be removed

# Extract hgnc-entrez comparison for duplicated entries
gdsc_hgnc_entrez_dups <- gdsc_hgnc_entrez_trunc[gdsc_hgnc_entrez_trunc$hgnc_symbol %in% dups, ]
gdsc_hgnc_entrez_dups_keep <- gdsc_hgnc_entrez_dups[!(gdsc_hgnc_entrez_dups$entrezgene_id 
                                                      %in% remove_genes), ]
gdsc_hgnc_entrez_dups_keep <- gdsc_hgnc_entrez_dups_keep[!(row.names(gdsc_hgnc_entrez_dups_keep) 
                                                           %in% remove_rows), ]
gdsc_hgnc_entrez_dups_remove <- gdsc_hgnc_entrez_dups[!row.names(gdsc_hgnc_entrez_dups) 
                                                      %in% row.names(gdsc_hgnc_entrez_dups_keep), ]
gdsc_hgnc_entrez_trunc <- gdsc_hgnc_entrez_trunc[!(gdsc_hgnc_entrez_trunc$entrezgene_id 
                                                   %in% gdsc_hgnc_entrez_dups_remove$entrezgene_id), ]


# Make changes in actual expression dataset
expr_gdsc_data <- expr_gdsc_data[expr_gdsc_data$GENE_SYMBOLS 
                                 %in% gdsc_hgnc_entrez_trunc$hgnc_symbol, ]
expr_gdsc_data <- expr_gdsc_data[order(expr_gdsc_data$GENE_SYMBOLS), ]
gdsc_hgnc_entrez_trunc <- gdsc_hgnc_entrez_trunc[order(gdsc_hgnc_entrez_trunc$hgnc_symbol), ]
row.names(expr_gdsc_data) <- gdsc_hgnc_entrez_trunc$entrezgene_id
expr_gdsc_rows <- row.names(expr_gdsc_data) # For use later when converting matrix --> df


#* Set colnames to cosmic id ---------------------------------------------------
expr_gdsc_data <- expr_gdsc_data[ , !(colnames(expr_gdsc_data) 
                                      %in% c("GENE_title", "GENE_ID", "GENE_SYMBOLS"))]
expr_gdsc_colnames <- colnames(expr_gdsc_data)
expr_gdsc_colnames <- strsplit(expr_gdsc_colnames, "DATA.")
expr_gdsc_cell_lines <- c()
for (n in expr_gdsc_colnames){
  expr_gdsc_cell_lines <- c(expr_gdsc_cell_lines, n[2])
}
colnames(expr_gdsc_data) <- expr_gdsc_cell_lines
expr_gdsc_data <- as.data.frame(sapply(expr_gdsc_data, as.numeric))
row.names(expr_gdsc_data) <- expr_gdsc_rows

# Remove rows that aren't found in meta data
expr_gdsc_data <- expr_gdsc_data[ , colnames(expr_gdsc_data) %in% cell_line_details$COSMIC.identifier]
```

# Save Data

And finally, we'll save all our data into a single .RData file.
```{r save_data}
save(dr_gdsc2, 
     expr_gdsc_data, 
     cell_line_details, 
     file = here("Data", "cleaned_gdsc_data.RData"))
# load(here("Data", "cleaned_gdsc_data.RData"))
```
